### 2012-10-17 rnaseq_paper.tex
more revisions

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome assembly, which is
lacking approximately 5\% of its true gene content (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

The above approaches recovered transcript fragments, but not entire genes
(figure in suppl?)
To construct a more comprehensive gene set containing all of the
assembled contigs, we tried using Oases-M to merge the assemblies from
multiple $k$ values \cite{Schulz:2012je}.  While it has been
demonstrated that merged transcripts from multiple-k assemblies
contain more isoforms than those from any single $k$, the sensitivity
of Oases-M for recovering splice variants has not been fully
evaluated.
% @CTB what does this mean? ``not fully evaluated''?
We merged transcripts from our global assembly, above,
with Oases-M using a k-mer size of 27, and compared them with the
unmerged transcripts.  We then cross-validated using publicly
available ESTs, which were not used in our assembly.
The results show that
Oases-M and the unmerged assembly share about 95\% of the predicted,
splice junctions, with 5\% disjoint.  Of these 5\%,
approximately 6\% of the Oases-M-specific splice junctions are
independently supported by ESTs, while 19\% of the unmerged splice
junctions are supported by ESTs.
% @CTB change from percent to total number, with percent in parantheses
This suggests that Oases-M probably
discards a number of real splice variants, although the unmerged
assembly is also missing some found by Oases-M.

@@

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared both methods to the ENSEMBL gene annotations,
which are produced by a pipeline that incorporates de novo gene
prediction and homology-based approaches as well as expression data.

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
66\% of splice junctions from ESTs and ENSENBL respectively, which is
10-15\% greater than that from corresponding unmerged gene models.
% @CTB what about splice junctions etc??

@@

To validate our predicted isoforms, we extracted coding sequences from
our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
of our isoforms from 12,399 distinct genes match mouse proteins at a
bit score of XXX.
Most of these matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few match, really?

@@

\subsection*{Leftover text}

@@

\section*{An overestimated number of genes and transcripts}

@@

\subsubsection{Cufflinks}

@@

To validate our predicted isoforms, we extracted coding sequences from
our gene models by using ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
of our isoforms from 12,399 genes match mouse proteins.
Most of these matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few, really?

\section*{An overestimated number of genes and transcripts}

The number of genes from our pipeline might be overestimated due to
fragmented transcripts.  In our datasets, read coverage is lower at
the $5'$ end due to the method used to convert mRNA to cDNA in library
preparation.  A bias of read coverage toward $3'$ end and low
expression level result in fragmented transcripts as shown in
Fig.~\ref{fragmented_transcripts}.  Moreover, spurious splice
junctions due to duplication in genome sequence can lead to a large
number of dubious isoforms.

>>>>

We next used Tophat to align mRNAseq reads to the genome and partition
reads by chromosome; we then assembled the partitioned reads with
Velvet and Oases using the same range of parameters as the global
assembly, above.  While the local assemblies were considerably more
computationally efficient, they lacked several thousand unique regions
that were present in the global assembly
% @CTB why are the global ones not included in the final gene models?
(Table~\ref{unique_sequences_matched_mouse}); this is probably due to
the incomplete nature of the current chicken genome assembly, which is
lacking approximately 5\% of its true gene content (cite).
Interestingly, over a hundred regions were present {\em only} in the
local assemblies, suggesting that the local assemblies might be
recovering additional exons.  Significant numbers of unique regions
from both global and local assemblies showed homology to the mouse
genome, indicating that at least some of these unique sequences
represented real sequence content.  Figure~\ref{global_vs_local} shows
an example of different isoforms detected by the two assembly methods.

@@

The above approaches recovered transcript fragments, but not entire genes
(figure in suppl?)
To construct a more comprehensive gene set containing all of the
assembled contigs, we tried using Oases-M to merge the assemblies from
multiple $k$ values \cite{Schulz:2012je}.  While it has been
demonstrated that merged transcripts from multiple-k assemblies
contain more isoforms than those from any single $k$, the sensitivity
of Oases-M for recovering splice variants has not been fully
evaluated.
% @CTB what does this mean? ``not fully evaluated''?
We merged transcripts from our global assembly, above,
with Oases-M using a k-mer size of 27, and compared them with the
unmerged transcripts.  We then cross-validated using publicly
available ESTs, which were not used in our assembly.
The results show that
Oases-M and the unmerged assembly share about 95\% of the predicted
splice junctions, with 5\% disjoint.  Of these 5\%,
approximately 6\% of the Oases-M-specific splice junctions are
independently supported by ESTs, while 19\% of the unmerged splice
junctions are supported by ESTs.
% @CTB change from percent to total number, with percent in parantheses
This suggests that Oases-M probably
discards a number of real splice variants, although the unmerged
assembly is also missing some found by Oases-M.

@@

% @CTB discuss: single ended, short reads => difficult to construct
% gene models.  How are isoforms assembled by oases, anyway!?

@@

We next compared the Gimme gene models to those produced by Cufflinks,
another reference-based approach to building gene models from mRNAseq
data\cite{Trapnell:2010kd}.  We also compared the results from both methods to the ENSEMBL gene annotations,
which are produced by a pipeline that incorporates de novo gene
prediction and homology-based approaches as well as expression data.

@@

As shown above, Cufflinks and the assembly method detect a number of
distinct but equally valid splice junctions, which suggests that we
could obtain greater sensitivity to exon-exon junctions in our gene models by merging both sets of predictions.
We therefore used Gimme to merge the Cufflinks and assembly gene
models Table.~\ref{genes_transcripts}.  This resulted in a decreased
number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{cufflinks_asm}). The merged gene models recover 51\% and
66\% of splice junctions from ESTs and ENSEMBL respectively, which is
10-15\% greater than that from corresponding unmerged gene models.
% @CTB what about splice junctions etc??

% @CTB Note that these results are, frankly, kind of stunning.  Discuss
% spleen/immune system diversity.

@@

To validate our predicted isoforms, we extracted putative coding sequences from
our gene models with ESTScan\cite{Iseli:1999vd}.  ESTScan
successfully translated 22,488 of 27,732 (81.1\%) of our isoforms to
protein sequences with 50 or more amino acids.  We then searched for
homologous sequences in mouse ENSEMBL, and found that 17,620 (78.4\%)
of our isoforms from 12,399 distinct genes match mouse proteins at a
bit score of XXX.
% @CTB
Most of these matches have a bit-score/length ratio greater than 1,
which indicates a good agreement between chicken and mouse proteins.
% @CTB why so few match, really?

@@

\section*{Leftover Results text}

@@

\subsection*{An overestimated number of genes and transcripts}

@@

@@
